---
title: "Examples demonstrating slcma R package v0.9"
author: Andrew Smith
date: 08 November 2024
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples demonstrating slcma R package v0.9}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputnc}
---

## Getting started

Load the library.

```{r}
library(slcma)
```

## Example 1. SLCMA

Generate some data that matches that seen in Mishra et al (2009).
This is the code supplied in Smith et al (2015).

```{r}
N <- c(141,20,88,80,40,35,317,367)
BMI <- c(28.7,27.5,29.1,27.8,28.3,27.1,27.6,26.0)
SE <- c(0.5,1.1,0.7,0.7,0.8,0.9,0.3,0.2)
S1 <- c(0,1,0,0,1,1,0,1)
S2 <- c(0,0,1,0,1,0,1,1)
S3 <- c(0,0,0,1,0,1,1,1)
x1 <- rep(S1,N)
x2 <- rep(S2,N)
x3 <- rep(S3,N)
epsilon <- rnorm(sum(N))
e <- lm(epsilon ~ x1 * x2 * x3)$residuals
y <- rep(BMI,N) + rep(SE*sqrt(N),N) * e/sd(e)
```

As in Smith et al (2015), perform SLCMA with 

* 3 critical periods (early, middle, late)

* Accumulation

* Early-middle mobility (up and down)

* Middle-late mobility (up and down)

## LARS (stage 1)

```{r}
myslcma <-
  slcma(y ~ x1 + x2 + x3 + Accumulation(x1,x2,x3) + Mobility(x1,x2) + Mobility(x2,x3))
```

Get a summary of the SLMCA, to see what variables are added at which stage.
This is more meaningful that the usual output of lars.

```{r}
summary(myslcma)
```

## Elbow plots

Default plot
```{r}
plot(myslcma)
```

For diagnostic purposes, 
can also show the part of the elbow plot after variables get removed and others added in.

```{r}
plot(myslcma, show.remove=TRUE, show.labels=FALSE)
```

Sometimes it is more helpful to show
an alternative elbow plot with the R2 from the relaxed lasso.

```{r}
plot(myslcma, relax=TRUE)
```

Can add your own labels to the elbow plot
```{r}
plot(myslcma, labels = c("Early critical period","Middle critical period","Late critical period",
                         "Accumulation","Early-middle mobility (up)","Early-middle mobility (down)",
                         "Middle-late mobility (up)", "Middle-late mobility (down)"))
```

Or if you can't remember all the variable names,
and just want to label the ones that appear on the plot.
```{r}
plot(myslcma, labels = c("Accumulation","Early critical period","Middle-late mobility (down)",
                         "Late critical period", "Early-middle mobility (up)"), selection.order=TRUE)
```

The function provides a sanity check so you can compare the default labels with the user-defined labels.

## Inference (stage 2)

Maybe there is an elbow at Step 3, perform selective inference by default
```{r}
slcmaInfer(myslcma, 3)
```

We have the option to report the results of the relaxed lasso instead
```{r}
slcmaInfer(myslcma, 3, method="relax")
```

We have the option of trying the max-|t| test, but only at Step 1.

```{r}
slcmaInfer(myslcma, 1, method="maxt")
```

We can also get confidence intervals from the max-|t| test, 
but these are time-consuming so they are not produced by default
```{r}
slcmaInfer(myslcma, 1, method="maxt", do.maxtCI=TRUE)
```

We have the option of Bayesian inference,
to find the posterior probability of the first-selected variable being the most important
```{r}
slcmaInfer(myslcma, 1, method="Bayes")
```

It is possible to produce results for all inference methods
```{r}
slcmaInfer(myslcma, 1, method=c("selectiveInference","relax","maxt","Bayes"))
```

## Example 2. Adjusting for covariates

Generate some data
(this is the code supplied in Smith et al (2016)).
```{r}
n <- 400
set.seed(1234)
covariate <- rnorm(n)
x1 <- covariate + rnorm(n)
x2 <- x1 + rnorm(n)
x3 <- x2 + 2*rnorm(n)
y <- 2*covariate + x3 + 3*rnorm(n)
```


## Adjust for covariates (stage 1)


Add any covariates into the formula,
then use the option adjust = a numeric vector containing the position(s) of covariate(s) in the formula.
Here there is one covariate (called 'covariate') and it is in the 8th position after the ~.
Hence adjust=8.

```{r}
slcmaAdjust <-
  slcma(y ~ x1 + x2 + x3 + Accumulation(x1,x2,x3) + Change(x1,x2) + Change(x2,x3) + Change(x1,x3) + covariate,
        adjust=8)
```

The function provides a sanity check so you can check the variables have the desired roles

You can revisit the sanity check
```{r}
slcmaAdjust$sanity
```

Can summarize and produce plots as before
```{r}
summary(slcmaAdjust) 
plot(slcmaAdjust)
```

Can perform inference (stage 2) as before
```{r}
slcmaInfer(slcmaAdjust,1, method=c("selectiveInference","relax","maxt","Bayes"))
```

## Example 3. Covariates and missing values

Some simulated data with missing values
(code originally supplied to The Dunn Lab in 2016)

```{r}
set.seed(123)
covariate <- rnorm(250)
x1 <- covariate + rnorm(250)
x2 <- x1 + 2*rnorm(250)
x3 <- x2 + 2*rnorm(250)
y <- 2*covariate + 2*(x3-x1) + 5*rnorm(250)
covariate[sample(1:250, size=10)] <- NA
x1[sample(1:250, size=10)] <- NA
x2[sample(1:250, size=10)] <- NA
x3[sample(1:250, size=10)] <- NA
y[sample(1:250, size=10)] <- NA
incompleteData <- data.frame(covariate,x1,x2,x3,y)
```

Complete case analysis
```{r}
CCslcma <- slcma(y ~ x1 + x2 + x3 + Accumulation(x1,x2,x3) + 
                 Change(x1,x3) + Ever(x1,x2,x3) + covariate, incompleteData, adjust=7)
```

Impute missing values (5 imputated datasets to save time)
```{r}
imputedData <- mice(incompleteData, m=5, maxit=20, seed=321)
```

Apply LARS to data with same covariance structure as pooled data across imputations 
 (data created via Cholesky decomposition)
 
```{r}
MIslcma <- slcma(y ~ x1 + x2 + x3 + Accumulation(x1,x2,x3) + 
                 Change(x1,x3) + Ever(x1,x2,x3) + covariate, data=imputedData, adjust=7)
```


Can summarize and produce plots as before
```{r}
summary(MIslcma) 
plot(MIslcma)
```

Can perform inference (stage 2) as before
```{r}
slcmaInfer(MIslcma,1, method=c("selectiveInference","relax","maxt","Bayes"))
```

Can recreate the dummy dataset used within the above SLCMA
```{r}
myMIdata <- MI2dummy(y ~ x1 + x2 + x3 + Accumulation(x1,x2,x3) + 
                     Change(x1,x3) + Ever(x1,x2,x3) + covariate, imputedData)
```
